---
output: 
  github_document: 
    pandoc_args: --webtex=https://latex.codecogs.com/svg.latex?
bibliography: ref.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  cache.path = "cache/README-",
  out.width = "100%", fig.width = 7, fig.height = 4
)
options(digits = 4)
```

# pedmod: Pedigree Models

The pedmod package provides functions to estimate models for pedigree data. 
Particularly, the package provides functions to estimate mixed models of 
the form:

$$\begin{align*}
Y_{ij} \mid \epsilon_{ij} = e 
  &\sim \text{Bin}(\Phi(\vec\beta^\top\vec x_{ij} + e), 1) \\
\vec\epsilon_i &= (\epsilon_{i1}, \dots, \epsilon_{in_i})^\top \sim
  N^{(n_i)}\left(\vec 0, \sum_{l = 1}^K\sigma_l^2 C_{il}
  \right)
\end{align*}$$

where $Y_{ij}$ is the binary outcome of interest for individual $j$ in 
family/cluster $i$, $\vec x_{ij}$ is the individual's known covariates, 
$\Phi$ is the standard normal distribution's CDF, and 
$\text{Bin}$ implies a binomial distribution such if 
$z\sim \text{Bin}(p, n)$ then the density of $z$ is:

$$
f(z) = \begin{pmatrix} n \\ z \end{pmatrix}p^zp^{n-z}
$$

The $C_{il}$s are known scale/correlation matrices where each of the $l$'th 
types correspond to a type of effect. A 
typical example is that $C_{il}$ is two times the kinship matrix in which 
case we call:

$$
\frac{\sigma_l^2}{1 + \sum_{k = 1}^K\sigma_k^2}
$$

the heritability. The scale parameters, the $\sigma_k^2$s, may be the 
primary interest in an analysis.

This package provides randomized quasi Monte Carlo methods to approximate 
the log marginal likelihood for these types of 
models with an arbitrary number scale matrices, $K$, and the derivatives 
with respect to $(\vec\beta^\top, 2\log\sigma_1,\dots, 2\log\sigma_K)^\top$.
We have re-written the Fortran code by @Genz02 in C++, made it easy to 
extend from a log marginal likelihood approximation to other approximations 
such as the derivatives, and added less precise but faster approximations 
of the $\Phi$ and $\Phi^{-1}$. Our own experience suggests that using the 
latter has a small effect on the precision of the result but can yield 
substantial reduction in computation times for moderate sized 
families/clusters.

The approximation by @Genz02 have already been used to estimate these 
types of models [@Pawitan04]. However, not having the gradients may slow 
down estimation substantially. Moreover, our implementation supports 
computation in parallel which is a major advantage given the 
availability of multi-core processors. 

Since the implementation is easy to extend, possible extensions are: 

 1. Survival times using mixed generalized survival models [@Liu17] with 
    a similar random effect structure as the model shown above. This way, 
    one avoids dichotomizing outcomes and can account for censoring.
 2. Generalized linear mixed model with binary, binomial, ordinal, or 
    multinomial outcomes with a probit link. The method we use here may be
    beneficial if the number of random effects per cluster is not much 
    smaller then the number observations in each cluster.
    
## Installation
The package can be installed from Github by calling:

```{r how_to_install, eval = FALSE}
remotes::install_github("boennecd/pedmod")
```

## Example

First, we source a file to get a function to simulate a data set with 
a maternal effect and a genetic effect like in @Mahjani20:

```{r source_sim_file}
# souce the file to get the simulation function
source(system.file("gen-pedigree-data.R", package = "pedmod"))

# simulate a data set
set.seed(68167102)
dat <- sim_pedigree_data(n_families = 1000L)

# distribution of family sizes
par(mar = c(5, 4, 1, 1))
plot(table(sapply(dat$sim_data, function(x) length(x$y))), 
     xlab = "Family size", ylab = "Number of families", bty = "l")
# total number of observations
sum(sapply(dat$sim_data, function(x) length(x$y)))

# average event rate
mean(unlist(sapply(dat$sim_data, `[[`, "y")))
```

As @Mahjani20, we have data families linked by three generations but we 
only have data for the last generation. We illustrate this with the first 
family in the simulated data set:

```{r one_family}
# this is the full family 
library(kinship2)
fam1 <- dat$sim_data[[1L]]
plot(fam1$pedAll)
# here is the C matrix for the genetic effect
rev_img <- function(x, ...)
  image(x[, NROW(x):1], ...)
cl <- colorRampPalette(c("Red", "White", "Blue"))(101)

par(mar = c(2, 2, 1, 1))
rev_img(fam1$rel_mat, xaxt = "n", yaxt = "n", col = cl, 
        zlim = c(-1, 1))
# the first part of the matrix is given below
with(fam1, rel_mat[seq_len(min(10, NROW(rel_mat))), 
                   seq_len(min(10, NROW(rel_mat)))])

# here is the C matrix for the maternal effect
rev_img(fam1$met_mat, xaxt = "n", yaxt = "n", col = cl, 
        zlim = c(-1, 1))
# each simulated family has such two matrices in addition to a design matrix
# for the fixed effects, X, and a vector with outcomes, y
str(fam1[c("X", "y")])
```

Then we perform the model estimation:

<!-- knitr::opts_knit$set(output.dir = ".") -->
<!-- knitr::load_cache("est_mod", path = "cache/README-") -->

```{r est_mod, cache=1}
# the true parameters are
dat$beta
dat$sc # the sigmas squared

# prepare the data to pass to the functions in the package
dat_arg <- lapply(dat$sim_data, function(x){
  # we need the following for each family: 
  #   y: the zero-one outcomes.
  #   X: the design matrix for the fixed effects. 
  #   scale_mats: list with the scale matrices for each type of effect.
  list(y = as.numeric(x$y), X = x$X,
       scale_mats = list(x$rel_mat, x$met_mat))
})

# create a C++ object
library(pedmod)
ll_terms <- get_pedigree_ll_terms(dat_arg, max_threads = 4L)

# get the starting values. This is very fast
y <- unlist(lapply(dat_arg, `[[`, "y"))
X <- do.call(rbind, lapply(dat_arg, `[[`, "X"))
start_fit <-  glm.fit(X, y, family = binomial("probit"))

# log-likelihood at the starting values without random effects
-sum(start_fit$deviance) / 2     
(beta <- start_fit$coefficients) # starting values for fixed effects 

# start at moderate sized scale parameters
sc <- rep(log(.2), 2)

# check log likelihood at the starting values. First we assign a function 
# to approximate the log likelihood and the gradient
fn <- function(par, seed = 1L, rel_eps = 1e-2, use_aprx = TRUE, 
               n_threads = 4L, indices = NULL, maxvls = 100000L){
  set.seed(seed)
  -eval_pedigree_ll(
    ll_terms, par = par, maxvls = maxvls, abs_eps = 0, rel_eps = rel_eps, 
    minvls = 1000L, use_aprx = use_aprx, n_threads = n_threads, 
    indices = indices)
}
gr <- function(par, seed = 1L, rel_eps = 1e-2, use_aprx = TRUE, 
               n_threads = 4L, indices = NULL, maxvls = 100000L){
  set.seed(seed)
  out <- -eval_pedigree_grad(
    ll_terms, par = par, maxvls = maxvls, abs_eps = 0, rel_eps = rel_eps, 
    minvls = 5000L, use_aprx = use_aprx, n_threads = n_threads, 
    indices = indices)
  structure(c(out), value = -attr(out, "logLik"), 
            n_fails = attr(out, "n_fails"))
}

# check output at the starting values
system.time(ll <- -fn(c(beta, sc)))
ll # the log likelihood at the starting values
system.time(gr_val <- gr(c(beta, sc)))
gr_val # the gradient at the starting values

# standard deviation of the approximation
sd(sapply(1:25, function(seed) fn(c(beta, sc), seed = seed)))

# we do the same for the gradient elements but only for a subset of the 
# log marginal likelihood elements
gr_hats <- sapply(1:25, function(seed) gr(c(beta, sc), seed = seed, 
                                          indices = 0:99))
apply(gr_hats, 1, sd)

# verify the gradient (may not be exactly equal due to MC error)
rbind(numDeriv = numDeriv::grad(fn, c(beta, sc), indices = 0:10, 
                                rel_eps = 1e-3, maxvls = 1e6), 
      pedmod   = gr(c(beta, sc), indices = 0:10, rel_eps = 1e-3, 
                    maxvls = 1e6))

# optimize the log likelihood approximation
system.time(opt <- optim(c(beta, sc), fn, gr, method = "BFGS"))
```

The output from the optimization is shown below:

```{r show_est_mod}
-opt$value      # the maximum log likelihood
opt$convergence # check convergence

# compare the estimated fixed effects with the true values
rbind(truth     = dat$beta, 
      estimated = head(opt$par, length(dat$beta)))

# compare estimated scale parameters with the true values
rbind(truth     = dat$sc, 
      estimated = exp(tail(opt$par, length(dat$sc))))
```

```{r svrg, eval = FALSE, echo = FALSE}
#####
# performs stochastic gradient descent instead (using SVRG).
#
# Args:
#   par: starting value.
#   gr: gradient function which return the function value as an attributate
#       called value.
#   n_clust: number of clusters.
#   batch_size: number of observations in each batch.
#   maxit: maximum number of iteration.
#   lr: learning rate.
#   verbose: print output during the estimation.
#   decay: numeric scalar used to decrease the learning rate.
#   ...: arguments passed to gr.
svrg <- function(par, gr, n_clust, batch_size, maxit = 10L, 
                 lr, verbose = FALSE, decay = .98, ...){
  indices <- sample.int(n_clust, replace = FALSE) - 1L
  blocks <- tapply(indices, (seq_along(indices) - 1L) %/% batch_size,
                   identity, simplify = FALSE)

  n_blocks <- length(blocks)
  n_par <- length(par)
  estimates <- matrix(NA_real_, n_par, maxit + 1L)
  fun_vals <- numeric(maxit + 1L)
  estimates[, 1L] <- par

  lr_use <- lr / decay
  V_mult <- qnorm(1 - .99 / maxit)
  for(k in 1:maxit + 1L){
    old_val <- estimates[, k - 1L]
    old_grs <- sapply(1:n_blocks - 1L, function(ii){
      idx_b <- (ii %% n_blocks) + 1L
      res_old <- gr(old_val, indices = blocks[[idx_b]], ...)
      c(attr(res_old, "value"), res_old)
    })

    fun_vals[k - 1L] <- sum(old_grs[1, ])
    old_grs <- old_grs[-1L, , drop = FALSE ]
    old_gr <- rowSums(old_grs) / n_blocks

    lr_use <- lr_use * decay
    for(ii in 1:n_blocks - 1L){
      idx_b <- (ii %% n_blocks) + 1L
      res <- gr(par, indices = blocks[[idx_b]], ...)
      fun_vals[k] <- fun_vals[k] + attr(res, "value")
      dir <- res - old_grs[, ii + 1L] + old_gr

      par <- par - lr_use * dir
    }

    estimates[, k] <- par

    if(verbose)
      cat(
        sprintf("End if iteration %4d with learning rate %.8f", k - 1L,
                lr_use),
        sprintf("Log marginal likelihood approximation is %12.2f", fun_vals[k]),
        sprintf("Previous approximate gradient norm was %14.2f\n",
                n_blocks * norm(as.matrix(old_gr))),
        sep = "\n")

    old_v <- fun_vals[k - 1L]
    new_v <- fun_vals[k]
  }

  list(result = par, fun_vals = fun_vals[2:k],
       estimates = estimates[, 2:k, drop = FALSE])
}

set.seed(1)
svrg_res <- svrg(c(beta, sc), gr = gr, n_clust = length(dat_arg), 
                 batch_size = 200L, lr = 2e-3, verbose = TRUE, maxit = 50L)
```

### Computation in Parallel

```{r pre_comp_par, echo = FALSE}
library(pedmod)
ll_terms <- get_pedigree_ll_terms(dat_arg, max_threads = 4L)
```

The method scales reasonably well in the number of threads as 
shown below:

```{r time_mult, cache = 1}
library(microbenchmark)
microbenchmark(
  `fn (1 thread)`  = fn(c(beta, sc), n_threads = 1),
  `fn (2 threads)` = fn(c(beta, sc), n_threads = 2),
  `fn (4 threads)` = fn(c(beta, sc), n_threads = 4),
  `gr (1 thread)`  = gr(c(beta, sc), n_threads = 1),
  `gr (2 threads)` = gr(c(beta, sc), n_threads = 2),
  `gr (4 threads)` = gr(c(beta, sc), n_threads = 4),
  times = 3)
```

### Using ADAM
We use stochastic gradient descent with the ADAM method [@Kingma15] in 
this section. We define a function below to apply ADAM and use it 
to estimate the model.

<!-- knitr::opts_knit$set(output.dir = ".") -->
<!-- knitr::load_cache("use_adam", path = "cache/README-") -->

```{r re_create_ptr, echo = FALSE}
library(pedmod)
ll_terms <- get_pedigree_ll_terms(dat_arg, max_threads = 4L)
```

```{r use_adam, cache = 1}
#####
# performs stochastic gradient descent (using ADAM).
#
# Args:
#   par: starting value.
#   gr: function to evaluate the log marginal likelihood.
#   n_clust: number of observation.
#   n_blocks: number of blocks.
#   maxit: maximum number of iteration.
#   seed: seed to use.
#   epsilon, alpha, beta_1, beta_2: ADAM parameters.
#   maxvls: maximum number of samples to draw in each iteration. Thus, it 
#           needs maxit elements.
#   verbose: print output during the estimation.
#   ...: arguments passed to gr.
adam <- function(par, gr, n_clust, n_blocks, maxit = 10L,
                 seed = 1L, epsilon = 1e-8, alpha = .001, beta_1 = .9,
                 beta_2 = .999, maxvls = rep(10000L, maxit), 
                 verbose = FALSE, ...){
  grp_dummy <- (seq_len(n_clust) - 1L) %% n_blocks
  n_par <- length(par)
  m <- v <- numeric(n_par)
  fun_vals <- numeric(maxit)
  estimates <- matrix(NA_real_, n_par, maxit)
  i <- -1L

  for(k in 1:maxit){
    # sample groups
    indices <- sample.int(n_clust, replace = FALSE) - 1L
    blocks <- tapply(indices, grp_dummy, identity, simplify = FALSE)
    
    for(ii in 1:n_blocks){
      i <- i + 1L
      idx_b <- (i %% n_blocks) + 1L
      m_old <- m
      v_old <- v
      res <- gr(par, indices = blocks[[idx_b]], maxvls = maxvls[k])
      fun_vals[(i %/% n_blocks) + 1L] <-
        fun_vals[(i %/% n_blocks) + 1L] + attr(res, "value")
      res <- c(res)

      m <- beta_1 * m_old + (1 - beta_1) * res
      v <- beta_2 * v_old + (1 - beta_2) * res^2

      m_hat <- m / (1 - beta_1^(i + 1))
      v_hat <- v / (1 - beta_2^(i + 1))

      par <- par - alpha * m_hat / (sqrt(v_hat) + epsilon)
    }
    
    if(verbose){
      cat(sprintf("Ended iteration %4d. Running estimate of the function value is: %14.2f\n", 
                  k, fun_vals[k]))
      cat("Parameter estimates are:\n")
      cat(capture.output(print(par)), sep = "\n")
      cat("\n")
    }

    estimates[, k] <- par
  }

  list(par = par, estimates = estimates, fun_vals = fun_vals)
}

#####
# use the function
# assign the maximum number of samples we will use
maxit <- 100L
minvls <- 250L
maxpts <- 10000L
maxpts_use <- exp(seq(log(2 * minvls), log(maxpts), length.out = maxit))

# show the maximum number of samples we use
par(mar = c(5, 4, 1, 1))
plot(maxpts_use, pch = 16, xlab = "Iteration number", bty = "l",
     ylab = "Maximum number of samples", ylim = range(0, maxpts_use))
set.seed(1)
system.time(
  adam_res <- adam(c(beta, sc), gr = gr, n_clust = length(dat_arg), 
                   n_blocks = 10L, alpha = 1e-2, maxit = maxit, 
                   verbose = FALSE, maxvls = maxpts_use, 
                   minvls = minvls))
```

The result is shown below.

```{r adam_re_create_ptr, echo = FALSE}
library(pedmod)
ll_terms <- get_pedigree_ll_terms(dat_arg, max_threads = 4L)
```

```{r res_adam}
-fn(adam_res$par) # the maximum log likelihood

# compare the estimated fixed effects with the true values
rbind(truth             = dat$beta,
      `estimated optim` = head(opt$par     , length(dat$beta)),
      `estimated ADAM`  = head(adam_res$par, length(dat$beta)))

# compare estimated scale parameters with the true values
rbind(truth             = dat$sc, 
      `estimated optim` = exp(tail(opt$par     , length(dat$sc))), 
      `estimated ADAM`  = exp(tail(adam_res$par, length(dat$sc))))

# could possibly have stopped much earlier maybe. Dashed lines are final 
# estimates
par(mar = c(5, 4, 1, 1))
matplot(t(adam_res$estimates), type = "l", col = "Black", lty = 1, 
        bty = "l", xlab = "Iteration", ylab = "Estimate")
for(s in adam_res$par)
  abline(h = s, lty = 2)
```

### The Multivariate Normal CDF Approximation

We compare the multivariate normal CDF approximation in this section 
with the approximation from the mvtnorm package which uses the implementation 
by @Genz02. The same algorithm is used 
but the version in this package is re-written in C++ and differs slightly.
Moreover, we have implemented an approximation of the standard normal CDF
and its inverse which reduces the computation time as we will show below.

```{r compare_w_mvtnorm, cache = 1}
#####
# settings for the simulation study
library(mvtnorm)
library(pedmod)
library(microbenchmark)
set.seed(78459126)
n <- 5L         # number of variables to interate out
rel_eps <- 1e-4 # the relative error to use

#####
# run the simulation study
sim_res <- replicate(expr = {
  # simulate covariance matrix and the upper bound
  S <- drop(rWishart(1L, 2 * n, diag(n) / 2 / n))
  u <- rnorm(n)
  
  # function to use pmvnorm
  use_mvtnorm <- function(rel_eps)
    pmvnorm(upper = u, sigma = S, algorithm = GenzBretz(
      abseps = 0, releps = rel_eps, maxpts = 1e7))
  
  # function to use this package
  use_mvndst <- function(use_aprx = FALSE)
    mvndst(lower = rep(-Inf, n), upper = u, mu = rep(0, n), 
           sigma = S, use_aprx = use_aprx, abs_eps = 0, rel_eps = rel_eps,
           maxvls = 1e7)

  # get a very precise estimate
  truth <- use_mvtnorm(rel_eps / 100)
  
  # computes the error with repeated approximations and compute the time it
  # takes
  n_rep <- 5L
  run_n_time <- function(expr){
    expr <- substitute(expr)
    ti <- get_nanotime()
    res <- replicate(n_rep, eval(expr))
    ti <- get_nanotime() - ti
    err <- (res - truth) / truth
    c(SE = sqrt(sum(err^2) / n_rep), time = ti / n_rep / 1e9)
  }
  
  mvtnorm_res        <- run_n_time(use_mvtnorm(rel_eps))
  mvndst_no_aprx_res <- run_n_time(use_mvndst(FALSE))
  mvndst_w_aprx_res  <- run_n_time(use_mvndst(TRUE))
  
  # return 
  rbind(mvtnorm            = mvtnorm_res, 
        `mvndst (no aprx)` = mvndst_no_aprx_res, 
        `mvndst (w/ aprx)` = mvndst_w_aprx_res)
}, n = 100, simplify = "array")
```

They have about the same average relative error as expected:

```{r show_averge_rel_err}
rowMeans(sim_res[, "SE", ])
par(mar = c(5, 4, 1, 1))
boxplot(t(sim_res[, "SE", ]))
# without extreme values
sum(keep <- colSums(sim_res[, "SE", ] > .5) < 1)
rowMeans(sim_res [, "SE", keep])
boxplot(t(sim_res[, "SE", keep]))
```

The new implementation is faster when the approximation is used:

```{r use_new_impl}
rowMeans(sim_res[, "time", ])
par(mar = c(5, 4, 1, 1))
boxplot(t(sim_res[, "time", ]))
```

## References
