---
title: Partitioning Families
bibliography: ../ref.bib
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, dpi = 45,
  comment = "#>",
  fig.path = "figures/partitioning-",
  cache.path = "cache/partitioning-",
  out.width = "100%", fig.width = 7, fig.height = 4,
  message = FALSE, warning = FALSE, error = FALSE)
options(digits = 4)
```

In many cases, the connected families that we have to integrate over may be 
large but sparsely connected. As an example, consider the following three 
generation family:

```{r pedigree_example}
# the data set we use
dat <- data.frame(
  id = 1:48, 
  mom = c(NA, NA, 2L, 2L, 2L, NA, NA, 7L, 7L, 7L, 3L, 3L, 3L, 3L, NA, 15L, 15L, 43L, 18L, NA, NA, 21L, 21L, 9L, 9L, 9L, 9L, NA, NA, 29L, 29L, 29L, 30L, 30L, NA, NA, 36L, 36L, 36L, 38L, 38L, NA, NA, 43L, 43L, 43L, 32L, 32L), 
  dad = c(NA, NA, 1L, 1L, 1L, NA, NA, 6L, 6L, 6L, 8L, 8L, 8L, 8L, NA, 4L, 4L, 42L, 5L, NA, NA, 20L, 20L, 22L, 22L, 22L, 22L, NA, NA, 28L, 28L, 28L, 23L, 23L, NA, NA, 35L, 35L, 35L, 31L, 31L, NA, NA, 42L, 42L, 42L, 45L, 45L), 
  sex = c(1L, 2L, 2L, 1L, 1L, 1L, 2L, 1L, 2L, 2L, 2L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 2L, 1L, 1L, 2L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 2L, 1L, 2L, 1L, 1L, 2L, 2L, 1L, 1L, 2L, 2L))

# plot the pedigree
library(kinship2, quietly = TRUE)
ped <- with(dat, pedigree(id = id, dadid = dad, momid = mom, sex = sex))
par(mar = c(1, 1, 1, 1))
plot(ped)
```

In the plot, circles are females and squares are males. An individual may be 
repeated in which case duplicate entries are illustrated by a dashed line. 
If we observe outcomes for all 48 members of the family then we have to 
do 48 dimensional as everybody is connected
under the assumption that the random effects that we 
specify are such that if individuals are assumed to be independent 
genetically then they are also independent with respect to any other random 
effect we may include (e.g. maternal effect or environmental effect). We assume 
that this is true in the rest of this vignette. Though it is possible also 
use the same method we describe here in other cases if one specifies a suitable 
graph, like the one we will get to soon, for the random effect structure.

To see that everybody is connected, notice that 

 - Child 16 is cousins with child 19 through her father. 
 - Child 19 is cousins with child 47 through his mom. 
 - Child 47 is cousins with child 33 through her mother. 
 - Child 33 is cousins with child 25 through her father. 
 - Child 25 is cousins with child 11 through his mother. At this point, we can 
   realize that we are able to get to every individual in the pedigree starting 
   at any point.
 - Child 11 is cousins with child 16 through her mother and this leads us back 
   to where we started and we have circle.
   
The relationships can also be represented as graph. For this purposes, we assign 
the functions below to create a list of edges between parents 
and their children. You can likely skip the functions. 

```{r assign_create_igraph_input}
# returns the mothers id.
# 
# Args:
#   pedigree: the pedigree object.
get_momid <- function(peddat)
  with(peddat, 
       vapply(mindex, function(x) if(x > 0L) id[x] else NA_integer_, 1L))

# returns the fathers id.
# 
# Args:
#   pedigree: the pedigree object.
get_dadid <- function(peddat)
  with(peddat, 
       vapply(findex, function(x) if(x > 0L) id[x] else NA_integer_, 1L))

# creates an edge list to pass to igraph. An edge is included between children 
# and parents.
# 
# Args:
#   pedigree: the pedigree object.
create_igraph_input <- function(peddat){
  id <- peddat$id
  father <- get_dadid(peddat)
  mother <- get_momid(peddat)
  
  # TODO: this is O(n^2)
  stopifnot(anyDuplicated(id) < 1)
  out <- lapply(id, function(x){
    # find the children
    children_idx <- which(x == father | x == mother)
    children <- if(length(children_idx) > 0)
      id[children_idx] else NULL
    
    # get the correct order (smallest first) and return
    is_larger <- x > children
    
    cbind(
      ifelse(is_larger, children, x        ), 
      ifelse(is_larger, x       , children))
  })
  
  out <- do.call(rbind, out)
  as.data.frame(out[!duplicated(out), ])
}
```

The graph version of the family looks like this:

```{r show_family_as_graph, message = FALSE}
# For some reason, kinship2::pedigree requires that we provide both a father 
# and mother or none. Therefore, we create a mock object. You can skip this
get_pedigree_mock <- function(id, dadid, momid, sex){
  if(is.factor(sex))
    sex <- as.integer(sex)
  
  # checks
  n <- length(id)
  stopifnot(n > 0, length(dadid) == n, length(momid) == n, length(sex) == n, 
            all(is.finite(sex)), all(sex %in% 1:2), 
            all(is.na(dadid) | dadid %in% id), 
            all(is.na(momid) | momid %in% id), 
            all(is.finite(id)))
  
  # create objects to return
  findex <- match(dadid, id, nomatch = 0L)
  mindex <- match(momid, id, nomatch = 0L)
  
  structure(
    list(famid = rep(1L, n), id = id, findex = findex, mindex = mindex,
         sex = factor(sex, levels = 1:2, labels = c("male", "famle"))), 
    class = "pedigree")
}

# assign function to plot the pedigree as a graph
do_graph_plot <- function(dat){
  ped <- with(dat, get_pedigree_mock(
    id = id, dadid = dad, momid = mom, sex = sex))
  library(igraph, quietly = TRUE)
  g_dat <- create_igraph_input(ped)
  graph_fam <-  graph.data.frame(g_dat, directed = FALSE)
  par(mar = c(1, 1, 1, 1))
  plot(graph_fam, vertex.size = 12, vertex.color = "gray",
       vertex.label.cex = .75, layout = layout_with_kk)
}
do_graph_plot(dat)
```

A node/vertex in the graph represents an individual and an edge indicates
a parent-child relation. The graph have the property that if we split 
(find a connected partition) 
the above into two sub-families (subgraphs) by removing child-parent relations
(removing edges) then we only have to do two smaller integrals rather than 
one large. Having said that, the 48 dimensional integral that we show here is 
not a problem but higher dimensional integrals may be.

## Implemented Methods

The above suggests the following procedure for simplifying the computational 
problem:

 1. Find a split (connected partition) such that the two sub-families 
    (subgraphs) are roughly the same size (an approximately balanced partition). 
 2. Of the sets that satisfies 1., find the one that removes the least amount
    of relationships (cuts the least amount of edges).

We have implemented procedures to do just this. It also turns out that it is 
fine to not satisfy number 1. That is, to not get two connected sets in the 
partition. We will return to this later and to what 
we have implemented but first provide an example. We start by splitting the 
above family in way that only satisfies 1.:

```{r only_balanced}
# get the partition 
library(pedmod)
only_balanced <- max_balanced_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L)
only_balanced$balance_criterion # the smallest of the two sets
removed_edges <- only_balanced$removed_edges
removed_edges # the relationships we have removed

# assign function to get the new data. You can likely skip this
get_reduced_data <- function(dat, removed_edges){
  for(i in 1:nrow(removed_edges)){
    . <- function(child, parent){
      idx   <- which(dat$id       == removed_edges[i, child])
      idx_m <- which(dat$mom[idx] == removed_edges[i, parent])
      if(length(idx_m > 0)){
        dat[idx[idx_m], "mom"] <<- NA_integer_
        return(TRUE)
      }
      idx_d <- which(dat$dad[idx] == removed_edges[i, parent])
      if(length(idx_d > 0)){
        dat[idx[idx_d], "dad"] <<- NA_integer_
        return(TRUE)
      }
      FALSE
    }
    if(.(1L, 2L))
      next
    .(2L, 1L)
  }
  
  dat
}
new_dat <- get_reduced_data(dat, removed_edges)

# redo the plot
do_graph_plot(new_dat)
```

The above shows the family after we remove the `r NROW(removed_edges)` 
relationships (edges) that our algorithm finds.
To illustrate where the cut is in the original graph,
we can color the vertices near the cut in the original graph:

```{r more_clear_only_balanced}
# assign function to show the split in the original graph
show_split <- function(dat, partition_obj){
  ped <- with(dat, pedigree(id = id, dadid = dad, momid = mom, sex = sex))
  g_dat <- create_igraph_input(ped)
  graph_fam <-  graph.data.frame(g_dat, directed = FALSE)
  nam <- vertex_attr(graph_fam)$name
  V(graph_fam)$color[nam %in% partition_obj$set_1] <- "lightblue" 
  V(graph_fam)$color[nam %in% partition_obj$set_2] <- "lightgray" 
  par(mar = c(1, 1, 1, 1))
  plot(graph_fam, vertex.size = 12, vertex.label.cex = .75, 
       layout = layout_with_kk)
}
show_split(dat, only_balanced)
```

### Reducing the Cut Cost

Next, we use the `slack` argument to reduce the number of relationships we 
remove (the number of edges we cut):

```{r also_reduce_cost}
also_cut <- max_balanced_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1)
also_cut$balance_criterion # the smallest of the two sets
removed_edges <- also_cut$removed_edges
removed_edges # the relationships we have removed

# redo the plot
show_split(dat, also_cut)
```

We only removed  `r NROW(removed_edges)` relationships (edges) this time.

### Using Individual Weights (Vertex Weights)
In many applications, we only observe some individuals. Therefore, we want the 
to split the family into equal sizes in terms of the individuals that we 
actually observe. We can do this by providing individual weights (vertex weights). 

As an example, we will take the family we have been working with and assume 
that we only observe individuals in the final generation. Thus, we will set the
weight of these individuals to a one and the remaining to some small positive 
number (say $10^{-5}$):

```{r use_weights}
# add the weights
is_final <- c(16:17, 11:14, 24:27, 33:34, 40:41, 47:48, 19L)
dat$id_weight <- ifelse(dat$id %in% is_final, 1., 1e-5)

weighted_partition <- max_balanced_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1, id_weight = dat$id_weight)
weighted_partition$balance_criterion # the smallest of the two sets
removed_edges <- weighted_partition$removed_edges
removed_edges # the relationships we have removed

# plot the new graph
plot_weighted <- function(dat, removed_edges){
  new_dat <- get_reduced_data(dat, removed_edges)
  ped <- with(new_dat, get_pedigree_mock(
    id = id, dadid = dad, momid = mom, sex = sex))
  g_dat <- create_igraph_input(ped)
  graph_fam <- graph.data.frame(g_dat, directed = FALSE)
  V(graph_fam)$color <- ifelse(vertex_attr(graph_fam)$name %in% is_final, 
                               "gray", "white")
  par(mar = c(1, 1, 1, 1))
  plot(graph_fam, vertex.size = 12, vertex.label.cex = .75, 
       layout = layout_with_kk)}

plot_weighted(dat, removed_edges)
show_split(dat, weighted_partition)
```

We have colored vertices which we observe in the first graph to make it easy 
to see that there is a as close as possible to an equal number in 
each sub-family. 

### Using Relationship Weights (Edge Weights)
We can notice that the previous solution removes relations between the 
individuals we observe and their immediate ancestors. We can add a larger weight
to these relationships to avoid or discourage this. To do so, we need to use the
`father_weight` and `mother_weight` arguments as shown below:

```{r use_edge_weights_too}
# add the weights
is_final <- c(16:17, 11:14, 24:27, 33:34, 40:41, 47:48, 19L)
dat$id_weight <- ifelse(dat$id %in% is_final, 1., 1e-5)

# add the edge weights
dat$father_weight <- dat$mother_weight <- ifelse(dat$id %in% is_final, 10., 1.)

# find the partition
weighted_partition <- max_balanced_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1, id_weight = dat$id_weight, father_weight = dat$father_weight, 
  mother_weight = dat$mother_weight)
weighted_partition$balance_criterion # the smallest of the two sets
removed_edges <- weighted_partition$removed_edges
removed_edges # the relationships we have removed

# plot the new graph
plot_weighted(dat, removed_edges)
show_split(dat, weighted_partition)
```

We end up cutting the link to the grandparents instead.

### Potentially Unconnected Partition

We may not require that the two sets in the partition is connected. The 
`unconnected_partition_pedigree` function does not impose this constraint. 
An example is given below:

```{r pedigree_unconnected}
# without weights
partition <- unconnected_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1)
partition$removed_edges # the relationships we have removed

# show the partition
show_split(dat, partition)
# with weights
weighted_partition <- unconnected_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1, id_weight = dat$id_weight, father_weight = dat$father_weight, 
  mother_weight = dat$mother_weight)
weighted_partition$removed_edges # the relationships we have removed

# show the partition
plot_weighted(dat, weighted_partition$removed_edges)
show_split(dat, weighted_partition)
```

We can also start from a connected partition like below:

```{r connected_start_pedigree_unconnected}
# get the connected partition
connected_partition <- max_balanced_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1, id_weight = dat$id_weight, father_weight = dat$father_weight, 
  mother_weight = dat$mother_weight)

# get the passably unconnected partition
partition <- unconnected_partition_pedigree(
  id = dat$id, father.id = dat$dad, mother.id = dat$mom, trace = 2L, 
  slack = .1, id_weight = dat$id_weight, father_weight = dat$father_weight, 
  mother_weight = dat$mother_weight, init = connected_partition$set_2)

# show the partition
show_split(dat, partition)
```

## Details

We will go through the details of the implemented methods in this section. 
All method we have implemented are for general graphs so we will will cast the 
problem for a general graph $G = (V, E)$ with vertices $V$ and edges $E$ 
with some vertex weight function $w$.
The first problem we need to solve is to implement a method to approximately 
solve the maximally balanced connected partition problem. That is, we want
to find a partition ($V_1$, $V_2$) of the graph such that $V_1$ is connected 
and $V_2$ is connected. The partition we find needs to maximize 
$\min (w(V_1), w(V_2))$. 

We have implemented the method suggested by 
@Chlebíková96. This methods requires that one can find the cut vertices 
(articulation points) and can construct the block-cut tree. 
Thus, we have implemented the method suggested by @Hopcroft73. The method 
suggested by @Hopcroft73 can be accessed through the 
`biconnected_components` and `block_cut_tree` function. 
We illustrate this below on a simulated graph:

```{r assign_sim_graph}
# simulates a connected graph of a given size
sim_graph <- function(size){
  stopifnot(size > 2)
  w_all <- (500:1)^(-2)
  out <- lapply(3:size, function(n){
    n_links <- sample.int(4L, 1L, prob = c(.5, .4, .05, .05))
    n_links <- min(n - 1L, n_links)
    idx_start <- max(1L, n - 500L)
    idx_end <- n - 1L
    prob <- if(n > 502L) w_all else (idx_end:idx_start)^(-2)
    cbind(from = n, to = sample(idx_start:idx_end, n_links, prob = prob))
  })
  out <- c(list(cbind(1L, 2L)), out)
  out <- do.call(rbind, out)
  ids <- sample.int(2L * size, size)
  out[] <- ids[out]
  out <- t(apply(out, 1L, sort))
  setNames(as.data.frame(out), c("from", "to"))
}
```

```{r simulate_small_graph}
# simulates a data set and illustrate it
set.seed(45L)
sim_dat <- sim_graph(50L)
par(mar = c(1, 1, 1, 1))
plot(graph.data.frame(sim_dat, directed = FALSE), layout = layout_with_kk,
     vertex.size = 12, vertex.color = "gray", vertex.label.cex = .75)
# get the biconnected components
cut_pts <- biconnected_components(from = sim_dat$from, to = sim_dat$to)
length(cut_pts) # number of biconnected components
```

The algorithm suggested by @Chlebíková96 can be used through the 
`max_balanced_partition` function with `slack == 0`:

```{r find_partition_simulate_small_graph}
# find the approximately balanced connected partition
max_part <- max_balanced_partition(from = sim_dat$from, to = sim_dat$to, 
                                   trace = 2L)
max_part$balance_criterion # the balance criterion 

# the removed edges
removed_edges <- max_part$removed_edges
removed_edges

# assign function to plot the partition
do_plot <- function(dat, partition_obj){
  removed_edges <- partition_obj$removed_edges
  
  to_remove <- apply(removed_edges, 1L, function(x)
    which(dat$from == x[1] & dat$to == x[2]))
  new_dat <- dat[-to_remove, ]
  par(mar = c(1, 1, 1, 1))
  plot(graph.data.frame(new_dat, directed = FALSE), layout = layout_with_kk,
       vertex.size = 12, vertex.color = "gray", vertex.label.cex = .75)
}
do_plot(sim_dat, max_part)
# assign function to show the split in the original graph
show_split <- function(dat, partition_obj, vertex.size = 12, 
                       vertex.label = NULL){
  graph_fam <-  graph.data.frame(dat, directed = FALSE)
  nam <- vertex_attr(graph_fam)$name
  V(graph_fam)$color[nam %in% partition_obj$set_1] <- "lightblue" 
  V(graph_fam)$color[nam %in% partition_obj$set_2] <- "lightgray" 
  par(mar = c(1, 1, 1, 1))
  plot(graph_fam, vertex.size = vertex.size, vertex.label.cex = .75, 
       layout = layout_with_kk, vertex.label = vertex.label)
}
show_split(sim_dat, max_part)
```

Our implementation of the method by @Chlebíková96 is outlined at
[cstheory.stackexchange.com/q/48864/62145](https://cstheory.stackexchange.com/q/48864/62145). 
Currently, we re-compute the cut vertices (articulation points) at every 
iteration but it may be possible to update these which presumably will be faster
and reduce the overall computation time. We do find a perfectly balanced 
partition 
above where each set is of size 25. However, we cut `r NROW(removed_edges)` edges
to this. Thus, we implemented a greedy method to reduce the number of cut edges
while maintaining a roughly balanced connected partition which we outline next.

Given an approximately balanced connected partition $(V_1, V_2)$ and vertex 
weight function $w_H$ as defined by @Chlebíková96, let 
$B$ be the block in which the partition is split,
$\tilde V_1 = V_1 \cap B$ and $\tilde V_2 = V_2 \cap B$, and let 
$I_1 \subseteq \tilde V_2$ be the vertices that are connected with any vertex 
in $\tilde V_1$ and define $I_2$ similarly. Then

 0. Compute $D$ as in the Kernighan–Lin algorithm [@Kernighan70].
 1. Clear the set of used vertices.
 2. For $i = 1,\dots,\min(\text{max moves},\text{vertices in }B)$:
    * Find the cut vertices $C_1$ of $\tilde V_1$ and $C_2$ of $\tilde V_2$.
    * Consider all the vertices in $I_1 \setminus C_2$ and $I_2 \setminus C_1$
      such that the balance criteria
      $w_H(V) / 2 - \epsilon w_H(V) \leq w_H(\tilde V_1)$ and 
      $w_H(\tilde V_1)\leq w_H(V) / 2 + \epsilon w_H(V)$ are 
      satisfied for some user defined $\epsilon \in [0, .5)$
      after the vertex is moved and that the vertex has 
      not yet been used.
    * If there are no vertices satisfying this then exit.
    * Select the vertex $v$ with the largest $D_v$. Break ties by 
      taking the $v$ with the largest balance criteria, 
      $\min (w_H(\lvert \tilde V_1\rvert), w_H(\lvert \tilde V_2\rvert))$. 
    * Let $g_i = D_v$ be the gain in this iteration, 
      move the vertex $v$, update $I_1$, $I_2$, and $D$. Add 
      $v$ to the used vertices.
 3. If no vertex were moved then exit. Otherwise find the 
    $l = \text{argmax}_{k = 1,\dots},\sum_{i = 1}^k g_i$. In case of ties, 
    take the one with the highest balance criterion.
 4. If $\sum_{i = 1}^l g_i \leq 0$ then undo all moves. Otherwise, undo moves 
    after move $l$ and repeat 1.
    
The method is very similar to the original Kernighan–Lin algorithm. 
It differs by the constraint that the two sets in the partition should be 
connected, that we only move one vertex in every move [like @Fiduccia82], 
and that the partition is only approximately balanced [like @Fiduccia82]. 
We also limit the number of moves we make as @Karypis98. Our implementations
to find a potentially unconnected partition 
(`unconnected_partition` and `unconnected_partition_pedigree`) 
proceeds in a similar manner as above except that we do not require that the
sets are connected. This substantially simplifies the algorithm.

$\epsilon$ is set with the `slack` argument of 
`max_balanced_partition` and the maximum number of moves to 
make in each iteration is set by the `max_kl_it_inner` argument of
`max_balanced_partition`. We illustrate the method below:

```{r partion_w_cut_simulate_small_graph}
# do the same but with also with the refinement step 
max_part <- max_balanced_partition(
  from = sim_dat$from, to = sim_dat$to, slack = .1, max_kl_it_inner = 50L, 
  max_kl_it = 1000L, trace = 1L)
max_part$balance_criterion # the balance criterion 
removed_edges <- max_part$removed_edges
removed_edges

# redo the plot
do_plot(sim_dat, max_part)
show_split(sim_dat, max_part)
```

### Larger Example
We will take a larger example where we also focus on the computation time. 
The example is shown below:

```{r larger_ex, fig.width=7, fig.height=7}
# simulate a data set with 500 observations
set.seed(45L)
sim_dat <- sim_graph(1000L)

# get the biconnected components
cut_pts <- biconnected_components(from = sim_dat$from, to = sim_dat$to)
length(cut_pts) # number of biconnected components

# find the approximately balanced connected partition
system.time(
  max_part <- max_balanced_partition(from = sim_dat$from, to = sim_dat$to, 
                                     trace = 1L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# improve on the number of cut vertices
system.time(
  max_part <- max_balanced_partition(from = sim_dat$from, to = sim_dat$to, 
                                     trace = 1L, slack = .1, 
                                     max_kl_it_inner = 1000L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# show the split
show_split(sim_dat, max_part, vertex.size = 3, vertex.label = NA)
```

We can also use the method for the potentially unconnected partition:

```{r unconnected_larger_ex, fig.width=7, fig.height=7}
# assign a variable with the connected partition
connected_partition <- max_part

# get the partition which may be unconnected
system.time(
  max_part <- unconnected_partition(from = sim_dat$from, to = sim_dat$to, 
                                    trace = 1L, slack = .1, 
                                    max_kl_it_inner = 1000L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# show the split
show_split(sim_dat, max_part, vertex.size = 3, vertex.label = NA)
# we can start with random initialization
init <- unique(c(sim_dat$from, sim_dat$to))
init <- sample(init, ceiling(length(init) / 2))

system.time(
  max_part <- unconnected_partition(from = sim_dat$from, to = sim_dat$to, 
                                    trace = 1L, slack = .1, 
                                    max_kl_it_inner = 1000L, 
                                    init = init))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# show the split
show_split(sim_dat, max_part, vertex.size = 3, vertex.label = NA)
# we can also start from the connected partition
system.time(
  max_part <- unconnected_partition(from = sim_dat$from, to = sim_dat$to, 
                                    trace = 1L, slack = .1, 
                                    max_kl_it_inner = 1000L, 
                                    init = connected_partition$set_1))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges
# compare with 
NROW(connected_partition$removed_edges)
show_split(sim_dat, max_part, vertex.size = 3, vertex.label = NA)
```

It is clear that the method is quite sensitive to the initialization.

### Even Larger Example

Here is an even larger example:

```{r largest_ex}
# simulate a data set with 100000 observations
set.seed(45L)
sim_dat <- sim_graph(100000L)

# get the biconnected components
cut_pts <- biconnected_components(from = sim_dat$from, to = sim_dat$to)
length(cut_pts) # number of biconnected components

# find the approximately balanced connected partition
system.time(
  max_part <- max_balanced_partition(from = sim_dat$from, to = sim_dat$to, 
                                     trace = 1L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# improve on the number of cut vertices
system.time(
  max_part <- max_balanced_partition(from = sim_dat$from, to = sim_dat$to, 
                                     trace = 1L, slack = .1, 
                                     max_kl_it_inner = 1000L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges
```

The computation time may depend on the ordering of the data set. It may be 
advantageous to allow our implementation to re-order the vertices e.g. when 
they are completely randomly ordered. We can see this my shuffling the data:

```{r shuffle_big_ex}
# shuffle the data
set.seed(1L)
sim_dat <- sim_dat[sample.int(NROW(sim_dat)), ]

# re-compute the partition without re-ordering the vertices
system.time(
  max_part_shuffled <- max_balanced_partition(
    from = sim_dat$from, to = sim_dat$to, trace = 1L, slack = .1, 
    max_kl_it_inner = 1000L, do_reorder = FALSE))
max_part_shuffled$balance_criterion # the balance criterion
NROW(max_part_shuffled$removed_edges) # number of removed edges

# re-compute the partition with re-ordering the vertices
system.time(
  max_part_shuffled_and_ordered <- max_balanced_partition(
    from = sim_dat$from, to = sim_dat$to, trace = 1L, slack = .1, 
    max_kl_it_inner = 1000L, do_reorder = TRUE))

# we got the same 
all.equal(max_part_shuffled_and_ordered, max_part_shuffled)

# may not match! 
all.equal(max_part, max_part_shuffled)
```

The solution will possibly not match when you reorder the data you pass to 
`max_balanced_partition`.

We can also compare the above with what we would have gotten if we had not 
required that the partition is connected.

```{r unconnected_largest_ex}
# assign a variable with the connected partition
connected_partition <- max_part_shuffled

# get potentially unconnected partition
system.time(
  max_part <- unconnected_partition(from = sim_dat$from, to = sim_dat$to, 
                                    trace = 1L, slack = .1, 
                                    max_kl_it_inner = 1000L))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges

# improve on the connected partition 
system.time(
  max_part <- unconnected_partition(from = sim_dat$from, to = sim_dat$to, 
                                    trace = 1L, slack = .1, 
                                    max_kl_it_inner = 1000L, 
                                    init = connected_partition$set_1))
max_part$balance_criterion # the balance criterion
NROW(max_part$removed_edges) # number of removed edges
```

The method completely failed this time except when we started from the 
balanced connected partition.

## References
